On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-10-13 Manuscript/manuscript.tex
rewrote the GLMM results

\jy{Add the results without 2022 to this table.}
\of{Does it make sense to order the table how I did, and block by dataset or
should I block by model?}
\begin{table}
  \centering
  \caption{AIC and estimated parameters for the three models fitted to the
    data.}
  \label{tab:Gamma_parameters}
  \begin{tabular}{c c c c c c c}
    \toprule
    Data set & Model & AIC & $\sigma_h$ & $\sigma_v$ & $\alpha$ & $\sqrt{\phi}$ \\
    \midrule
    Excluding 2022 &
      Venue Effect Only & $-1862.2$ &   & 0.048 & $-1.857$ & 0.152\\
     & Heat Effect Only & $-1910.9$ & 0.069 &  & $-1.877$ & 0.136\\
     & Venue and Heat Effect & $-1912.9$ & 0.058 & 0.033 & $-1.880$ & 0.136\\[1ex]
    Including 2022 &
    Venue Effect Only & $-2023.5$ &   & 0.054 & $-1.871$ & 0.150\\
    & Heat Effect Only & $-2070.7$ & 0.073 &  & $-1.890$ & 0.135\\
    & Venue and Heat Effect & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
    \bottomrule
  \end{tabular}
\end{table}

\jy{Hopefully, the full model is preferred for both with and without 2022}
\of{It is but the AIC are more similar without 2022}

The estimated parameters of the GLMM model from our analysis are shown in 
Table~\ref{tab:Gamma_parameters}. These statistics reflect the inclusion of 2022
in our analysis and shows the three models we considered. The standard deviation
of the heat effect is $\sigma_h$, the standard deviation of the venue effect
is $\sigma_v$, the intercept is $\alpha$ and we report $\sqrt{\phi}$ as the
square root of the dispersion parameter $\phi$.
The GLMM with both venue and heat random effects was the best model as evidenced
by a decrease in the AIC compared to the other models with only one random
effect.  When we exclude 2022 from the full model, we observe small but intriguing
discrepancies in the models. The standard deviation of the venue effect, 
$\sigma_v$, decreases from 0.40 to 0.33 when 2022 is excluded.  Thus, 2022
increased the noise in the overall data, but many of the other statistics
remained relatively similar, such as the standard deviation of the heat effect
and the dispersion parameter.

%\of{Can I just refer to it as the dispersion parameter or should I say "square
%root of the dispersion parameter" Mathematically it shouldn't make a difference 
%right as 0.134^2 and 0.136^2 are very similar and I am applying a function to 
%both? Also should I be saying "standard deviation of heat effect", or sigma_v 
%or both?}

\jy{discuss the differences in the parameter estimates with and without 2022
  data, especially the sd of the random effect of venue.}

Figure~\ref{fig:VenueEffects} shows the venue effects from 1999 to 2023 based
on the full model both with and without 2022.  As this graph does not show heat
\jy{discuss the years the venue effects are high and low}
effects, the estimates may not align exactly with the box plot presented in
figure~\ref{fig:Boxplot}.  The venue effect for 2022 is lower than any other
effect, but may not be as extreme as one would think considering the history
of reaction times going back to 1999.

Thus, results from the GLMM with both venue and heat random effects will be 
presented throughout the rest of the paper.

\begin{figure}[tbp]
  \centering
  \includegraphics{ComparisonOfVenueEffects}
  \caption{The venue effects from 1999 to 2023 for the data including 2022 (top)
  and excluding 2022 (bottom)}
  \label{fig:VenueEffects}
\end{figure}

Two aims of this section are to
determine how extreme a reaction time below 0.1 seconds is, and although the
venue effects did not differ significantly, to determine how the exclusion of 
2022 could alter the results. Using the gamma family with \texttt{link = log},
the probability of the reaction time being less than 0.1 seconds can be 
calculated in R using the methods described in Section~\ref{sec:glmm}. A 
simulation size of $10,000,000$ was used to estimate the probability of 
observing a reaction time below two thresholds: 0.1 and 0.08.

@@

Table~\ref{tab:Sim_probability} shows that changing the reaction time 
barrier from 0.1 seconds to 0.08 seconds drastically reduces the chances of 
observing a reaction time that would break the barrier. If World Athletics were
to change the reaction time the way that \citet{komi2009iaaf} suggests, the 
probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification. When including 2022,
a probability of $4.34\cdot10^{-3}$ corresponds to roughly 1 in every 230 starts
being below 0.1 seconds. When we exclude 2022, the probability of observing a
reaction time below 0.1 seconds decreased to $3.54\cdot10^{-3}$, or 1 in every
283 starts.  While this seems low, it is not unreasonable to think that certain
athletes with strong reaction times could begin approach the 0.1 second barrier,
especially when considering Allen's other reaction times: 0.124 and 0.101 at the
2022 World Track and Field Championships.  Another important conclusion from our
results is how excluding 2022 lowered the probability of observing a fast 
reaction time, suggesting that 2022 was peculiarly fast, which aligns with the
results presented in section~\ref{subsec:Results_Rank}.

@@

We can use the same model but alter our interpretation slightly to select a 
reaction time barrier given the probability of observing a time less than the 
barrier. When considering all the data, if we want a 0.1\% chance of 
observing an illegally fast reaction time, the reaction
time barrier should be 0.092 seconds, as shown in Table~\ref{tab:Sim_time}. We
can repeat this calculation for various probability levels,
for example a probability of 0.0001 (one disqualification for every 10,000 starts)
results in a reaction time barrier of 0.083 seconds. As the exclusion of 2022
from the data resulted in a lower probability of observing a low reaction time,
the reaction time barrier will be less lenient when calculating the barrier
based on probability.

@@

This paper does not consider data from the women's 100 meter hurdle, and although
that was initially because of concerns over reaction time differences in men and
women, exploratory analysis showed that 2022 was at best only
slightly faster relative to other championships going back to 2003.
Likewise, the rank-based comparisons resulted in p-values
that were not drastically significant after adjusting for multiple tests.
The lack of evidence from the women's data makes it 
harder to conclude that the sensors used at the 2022 championships were faulty, 
although it is unknown if the sensors for the men's 110 meter hurdle and women's
100 meter hurdle were the same. On the other hand, this may be related to 
another debated question regarding whether the reaction time barrier should be 
the same for men and women.

>>>>

\begin{table}
  \centering
  \caption{AIC and estimated parameters for the three models fitted to the
    data.}
  \label{tab:Gamma_parameters}
  \begin{tabular}{c c c c c c c}
    \toprule
    Data set & Model & AIC & $\sigma_h$ & $\sigma_v$ & $\alpha$ & $\sqrt{\phi}$ \\
    \midrule
    Excluding 2022 &
      Venue Effect Only & $-1862.2$ &   & 0.048 & $-1.857$ & 0.152\\
     & Heat Effect Only & $-1910.9$ & 0.069 &  & $-1.877$ & 0.136\\
     & Venue and Heat Effect & $-1912.9$ & 0.058 & 0.033 & $-1.880$ & 0.136\\[1ex]
    Including 2022 &
    Venue Effect Only & $-2023.5$ &   & 0.054 & $-1.871$ & 0.150\\
    & Heat Effect Only & $-2070.7$ & 0.073 &  & $-1.890$ & 0.135\\
    & Venue and Heat Effect & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
    \bottomrule
  \end{tabular}
\end{table}

The estimated parameters of the GLMM model under the log link are shown in 
Table~\ref{tab:Gamma_parameters}, with both including and excluding 2022 data.
Three models were considered, differing in there random effects: venue effect
only; heat effect only, and both venue and heat effects. Reported are the AIC
and estimated model parameters: the standard deviation
of the heat effect $\sigma_h$, the standard deviation of the venue
effect~$\sigma_v$, the intercept~$\alpha$, and the
square root of the dispersion parameter $\phi$. With or without 2022,
the GLMM with both venue and heat random effects was the best model as evidenced
by the AIC compared to the other models with only one random
effect.  When we exclude 2022 from the full model, we observe small but intriguing
discrepancies in the models. The standard deviation of the venue effect, 
$\sigma_v$, increases from 0.33 to 0.40 when 2022 is included, however, the
changes to the estimates of other parameters
remained relatively similar, such as the standard deviation of the heat effect
and the dispersion parameter.

\begin{figure}[tbp]
  \centering
  \includegraphics{ComparisonOfVenueEffects}
  \caption{The venue effects from 1999 to 2023 estimated from the GLMM for the
    data including 2022 (top) and excluding 2022 (bottom).}
  \label{fig:VenueEffects}
\end{figure}

Figure~\ref{fig:VenueEffects} further illustrates the venue effect from 1999 to
2023 based on the full GLMM both with and without 2022.  The temporal patterns
of the two sets of effects look very similar, except that the positive effects
estimated with 2022 included have higher magnitude than those estimated without
2022, which compensates the negative effect of the 2022 with a large magnitude.
Both plots agree to some extent with the pattern of the centers of the box plots in
Figure~\ref{fig:Boxplot}. The mismatch is due to the venue effect, which has a
higher standard deviation than the venue effect. The venue effect for 2022 is
lower than that for any other years, but may not be as extreme as one would
think considering the history of reaction times going back to 1999.

@@

The GLMM with both venue and heat random effects helps us to assess how extreme
a reaction time below a given threshold is. The probability of observing a reaction
time below a threshold, assuming no intentional false starts, can be
approximated by generating a large number of realizations from the fitted
GLMM. We used 10 millions realizations in the analysis. 
Table~\ref{tab:Sim_probability} summarizes the estimated probabilities of
observing a reaction time below 0.08, 0.09, and 0.10 seconds under two different
scenarios: one excluding and the other including data from 2022. Apparently,
excluding 2022 lowered the probability of observing a fast reaction time, but
the difference is not in a any big magtinude. For example, the probability for
seeing a reaction time below 0.10 seconds changes from $3.54\cdot 10^{-3}$ to
$4.20\cdot 10^{-3}$ when 2022 is included. They corresponding to roughly one in
every 230 starts and 283 starts, respectively. The table shows that changing the
reaction time barrier from 0.1 seconds to 0.08 seconds drastically reduces the
chance of observing a reaction time below the barrier, from one in roughly every
283 starts to one in every 29851 starts using data with 2022 included. The
results substantiates the recommendations by \citet{komi2009iaaf}.

@@

Utilizing the same model, we can subtly shift our interpretation to determine a
suitable reaction time barrier based on the probability of observing a time
below said barrier. Including 2022 data, a reaction time barrier of 0.092
seconds is warranted to maintain a 0.1\% chance of observing an illicitly fast
reaction time, as delineated in Table~\ref{tab:Sim_time}. This calculation can
be replicated for various probability levels; for instance, a probability of
0.0001 (equating to one disqualification per 10,000 starts) necessitates a
reaction time barrier of 0.083 seconds. The exclusion of 2022 data, which
inherently lowers the probability of observing a swift reaction time, mandates a
less lenient reaction time barrier when calculated based on probability, albeit
without a substantial difference in magnitude.

@@

We did not consider data from the women's 100 meter hurdle in the analysis.
In addition to the concerns over reaction time differences in men and
women, exploratory analysis showed that 2022 was at best only
slightly faster relative to other championships going back to 2003.
Likewise, the rank-based comparisons resulted in p-values
that were not drastically significant after adjusting for multiple tests.
The lack of evidence from the women's data makes it 
harder to conclude that the sensors used at the 2022 championships were faulty, 
although it is unknown if the sensors for the men's 110 meter hurdle and women's
100 meter hurdle were the same. On the other hand, this may be related to 
another debated question regarding whether the reaction time barrier should be 
the same for men and women.
